---
title: "serosurvey: Serological Surveys and Prevalence Estimation Under Misclassification"
# subtitle: "⚔<br/>with xaringan"
author: "Andree Valle-Campos (@avallecam)"
institute: "Centro Nacional de Epidemiología, CDC Perú"
# date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["useR", "useR-fonts", "custom.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
# Disable scrolling through slides to
# allow scrolling in slides
        scroll: false
---

class: center, middle

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE,tidyverse.quiet = TRUE)
knitr::opts_chunk$set(fig.width=10, fig.height=4)
# options(tidyverse.quiet = TRUE)
```

```{css, echo=FALSE}
pre {
  max-width: 200%;
  overflow-x: scroll;
}
```

# slides:<br/>[bit.ly/serosurvey-user21](https://bit.ly/serosurvey-user21)

---

# Introduction (1/2)

--

- Population-based serological surveys __generate__ seroprevalence estimates and __quantify__ how many people have been infected by a certain pathogen.

--

- These studies need to estimate prevalence considering __misclassification__ due to imperfect diagnostic tests with sensitivity and specificity __known with certainty__.

--

- However, during __first months__ of a novel pathogen outbreak, __performance__ of diagnostic tests are __“unknown” or “uncertain”__ given limited validation studies.

--

.footnote[
[1] Takahashi, S., et al. 2020. Are SARS-CoV-2 seroprevalence estimates biased?. _The Journal of Infectious Diseases_. https://doi.org/10.1093/infdis/jiaa523

[2] Flor, M., et al. 2020. “Comparison of Bayesian and Frequentist Methods for Prevalence Estimation Under Misclassification.” BMC Public Health 20 (1). https://doi.org/10.1186/s12889-020-09177-4.

[3] Kritsotakis, EI., 2020. “On the Importance of Population-Based Serological Surveys of SARS-CoV-2 Without Overlooking Their Inherent Uncertainties.” _Public Health in Practice_ 1 (November): 100013. [10.1016/j.puhip.2020.100013](https://doi.org/10.1016/j.puhip.2020.100013).
]

---

# Introduction (2/2)

--

- This situation opened the opportunity for the development of newer __ad-hoc methods__. Most of them used R, but not stored in a __R package__.

--

- The __increased demand__ for prevalence estimates and the absence of standardized procedures favored the usage of __non-reproducible workflows__.

--

- We created the __serosurvey R package__ to gather Serological Survey Analysis (known and recently developed) __functions__ and __workflow__ templates.

--

.footnote[
[1] Larremore, DB., Bailey K. Fosdick, Sam Zhang, and Yonatan H. Grad. 2020. “Jointly Modeling Prevalence, Sensitivity and Specificity for Optimal Sample Allocation.” _bioRxiv_, May. https://doi.org/10.1101/2020.05.23.112649.

[2] Gelman, Andrew, and Bob Carpenter. 2020. “Bayesian Analysis of Tests with Unknown Specificity and Sensitivity.” _Journal of the Royal Statistical Society: Series C (Applied Statistics)_, August. https://doi.org/10.1111/rssc.12435.

[3] Andree Valle Campos. 2020. avallecam/serosurvey: Serological Survey Analysis For Prevalence Estimation Under Misclassification (Version v1.0.1). Zenodo. http://doi.org/10.5281/zenodo.4118710. https://avallecam.github.io/serosurvey
]


---

class: chapter-slide

# The `serosurvey` functions

```{r example,echo=FALSE,warning=FALSE,message=FALSE}
library(serosurvey)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# additional
library(tidyverse)
library(srvyr)
library(survey)
library(tictoc)
library(furrr)
library(purrr)
# theme
theme_set(theme_bw())
```

```{r,echo=FALSE}
data(api)

datasurvey <- apiclus2 %>% 
  mutate(survey_all="survey_all") %>% 
  # create variables
  mutate(outcome_one = awards,
         outcome_two = cut(pct.resp,breaks = 2),
         covariate_01 = stype,
         covariate_02 = both)
```

```{r,echo=FALSE}
# tratamiento de stratos con un solo conglomerado
options(survey.lonely.psu = "certainty")

# uu_clean_data %>% count(CONGLOMERADO,VIVIENDA)

# diseño muestral de la encuesta ---------------------------------

design <- datasurvey %>% 
  
  filter(!is.na(outcome_one)) %>% #CRITICAL! ON OUTCOME
  filter(!is.na(pw)) %>% #NO DEBEN DE HABER CONGLOMERADOS SIN WEIGHT
  
  as_survey_design(
    id=c(dnum, snum), #~dnum+snum, # primary secondary sampling unit
    # strata = strata, #clusters need to be nested in the strata
    weights = pw # factores de expancion
  )
```

```{r,echo=FALSE}
# denominadores
covariate_set01 <- datasurvey %>% 
  select(covariate_01,
         #sch.wide,
         #comp.imp,
         covariate_02) %>% 
  colnames()

# numerators within outcome
covariate_set02 <- datasurvey %>% 
  select(#stype,
         #sch.wide,
         #comp.imp,
         covariate_02) %>% 
  colnames()
```

---

class: scrollable-slide

<!-- `survey` -->

# Estimate single prevalence

--

From a [`srvyr`](http://gdfe.co/srvyr/) [`survey`](https://r-survey.r-forge.r-project.org/survey/) design object, __`serosvy_proportion`__ estimates:

  + weighted prevalence (`prop`), 
  + total population (`total`),
  + raw proportion (`raw_prop`), 
  + coefficient of variability (`cv`), 
  + design effect (`deff`)

--

```{r}
serosvy_proportion(design = design, #<<
                   denominator = covariate_01,
                   numerator = outcome_one)
```

```{r,eval=FALSE,echo=FALSE}
example("serosvy_proportion")
```

---

<!-- `serology` -->

# Under misclassification

--

- We gather __one frequentist approach__ (Rogan and Gladen, 1978), 
available in Github repos (Azman et al. 2020; Takahashi et al. 2020).
Check [here](https://avallecam.github.io/serosurvey/reference/index.html).

--

- We provide __tidy outputs for bayesian approaches__ developed 
by Larremore, et al. for [known](https://github.com/LarremoreLab/covid_serological_sampling/blob/master/codebase/seroprevalence.R) and [unknown](https://github.com/LarremoreLab/bayesian-joint-prev-se-sp/blob/master/singleSERO_uncertainTEST.R) 
test performance.

--

- You can use them with [`purrr`](https://purrr.tidyverse.org/) and [`furrr`](https://davisvaughan.github.io/furrr/) to efficiently iterate 
and parallelize this step for __multiple prevalences__.
Check the workflow 
in [Article tab](https://avallecam.github.io/serosurvey/articles/howto-reprex.html).

--

.footnote[
[1] Rogan, Walter J., and Beth Gladen. 1978. “Estimating Prevalence from the Results of A Screening Test.” _American Journal of Epidemiology_ 107 (1): 71–76. https://doi.org/10.1093/oxfordjournals.aje.a112510.

[2] Larremore, DB, et al. 2021. “Estimating SARS-CoV-2 Seroprevalence and Epidemiological Parameters with Uncertainty from Serological Surveys.” _eLife_, March. https://doi.org/10.7554/eLife.64206.

[3] Larremore, DB., et al. 2020. “Jointly Modeling Prevalence, Sensitivity and Specificity for Optimal Sample Allocation.” _bioRxiv_, May. https://doi.org/10.1101/2020.05.23.112649.

]

---

# Known test performance

--

```{r,eval=FALSE}
serosvy_known_sample_posterior( #<<
  #in population
  positive_number_test = 321,
  total_number_test = 321+1234,
  # known performance
  sensitivity = 0.93,
  specificity = 0.975)
```

```{r,echo=FALSE,fig.alt="One ggplot panel with density as vertical y-axis. Histogram and density plot of the Posterior Distribution of the Prevalence accounting for misclassification from a diagnostic test with known performance. Additionally, it includes vertical lines at the mean, 5% and 95% percentiles."}
tidy_result <- serosvy_known_sample_posterior(
  #in population
  positive_number_test = 321,
  total_number_test = 321+1234,
  # known performance
  sensitivity = 0.93,
  specificity = 0.975
)

tidy_result_out <-
  tidy_result %>%
  select(summary) %>%
  unnest(cols = c(summary))

tidy_result %>%
 select(posterior) %>%
 unnest(cols = c(posterior)) %>%
 ggplot(aes(x = r1)) +
 geom_histogram(aes(y=..density..),binwidth = 0.0005) +
 geom_density() +
 geom_vline(aes(xintercept=tidy_result_out %>%
                  pull(numeric.mean)),
            color="red",lwd=1) +
 geom_vline(aes(xintercept=tidy_result_out %>%
                  pull(numeric.p05)),
            color="red") +
 geom_vline(aes(xintercept=tidy_result_out %>%
                  pull(numeric.p95)),
            color="red") +
 scale_x_continuous(breaks = scales::pretty_breaks())
```

```{r,eval=FALSE,echo=FALSE}
example("serosvy_known_sample_posterior")
```

---

# Unknown test performance

--

- When test 
sensitivity and specificity are not known with 
certainty.

--

```{r,eval=FALSE,echo=FALSE}
# result_unk <- sample_posterior_r_mcmc_testun(
#   samps = 10000,
#   #in population
#   pos = 692, #positive
#   n = 3212, #total
#   # in lab (local validation study)
#   tp = 670,tn = 640,fp = 202,fn = 74)
```

```{r,eval=FALSE}
serosvy_unknown_sample_posterior( #<<
  #in population
  positive_number_test = 321,
  total_number_test = 321+1234,
  # in lab (local validation study)
  true_positive = 670,
  true_negative = 640,
  false_positive = 202,
  false_negative = 74)
```

```{r,echo=FALSE,fig.alt="Three ggplot facet panels with the same vertical y-axis of density. Histogram and density plots of the Posterior Distribution of the (i) r: Prevalence accounting for misclassification from a diagnostic test with unknown performance; (ii) se: Sensitivity; (iii) sp: Specificity."}
result_unk <- serosvy_unknown_sample_posterior_ii(
  #in population
  positive_number_test = 321,
  total_number_test = 321+1234,
  # in lab (local validation study)
  true_positive = 670,
  true_negative = 640,
  false_positive = 202,
  false_negative = 74)

result_unk %>%
  select(posterior) %>% 
  unnest(posterior) %>%
  rownames_to_column() %>%
  pivot_longer(cols = -rowname,
               names_to = "estimates",
               values_to = "values") %>%
  ggplot(aes(x = values)) +
  geom_histogram(aes(y=..density..),binwidth = 0.0005) +
  geom_density() +
  facet_grid(~estimates,scales = "free_x")
```

```{r,eval=FALSE,echo=FALSE}
example("serosvy_unknown_sample_posterior")
```


---

class: chapter-slide

# The `serosurvey` workflow

---

# Multiple prevalences

--

- In 
the [Article tab](https://avallecam.github.io/serosurvey/articles/howto-reprex.html) 
we provide a workflow to __estimate multiple prevalences__: 

  + different sets of covariates and outcomes as numerators or denominators,
  
  + in one single pipe operation

--

```{r}
# set 01 of denominator-numerator
expand_grid( #<<
  design=list(design),
  denominator=c("covariate_01","covariate_02"), # covariates
  numerator=c("outcome_one","outcome_two") # outcomes
  ) %>% 
  # set 02 of denominator-numerator (e.g. within main outcome)
  union_all( #<<
    expand_grid(
      design=list(design),
      denominator=c("outcome_one","outcome_two"), # outcomes
      numerator=c("covariate_02") # covariates
    )) %>%
  # create symbols (to be read as arguments) 
  mutate(
    denominator=map(denominator,dplyr::sym), #<<
    numerator=map(numerator,dplyr::sym)
  ) %>%
  # estimate prevalence
  mutate(output=pmap(.l = select(.,design,denominator,numerator),
                     .f = serosvy_proportion)) %>% #<<
  # show the outcome
  select(-design,-denominator,-numerator) %>% 
  unnest(cols = c(output)) %>% 
  print(n=Inf)
```

---

# Add test performance

--

- Here we use `purrr` and `furrr` within a `for-loop` to efficiently update prevalence posterior distributions for multiple prevalences.

--

- We apply the procedure to each point estimate and confidence interval

--

- Watch the results in 
the [Article tab](https://avallecam.github.io/serosurvey/articles/howto-reprex.html) 

--

```{r,eval=FALSE,echo=TRUE}
# 56-60sec per covariate
# 4GB RAM
# parallelized in 8 cores using purrr y furrr

plan(multisession, workers = availableCores()) #<<
tic()

out <- tibble()

for (i in 1:nrow(outcome_01_adj_pre)) { #<<
  
  out <- outcome_01_adj_pre %>% 
    
    slice(i) %>% 
    
    # dot estimate
    mutate(adj_dot_unk=future_pmap(
      .l = select(.,
                  positive_number_test=total_round,
                  total_number_test=total_den_round),
      .f = possibly(serosvy_unknown_sample_posterior,otherwise = NA_real_), #<<
      true_positive = true_positive,
      false_negative = false_negative,
      false_positive = false_positive,
      true_negative = true_negative)) %>% 
    serosvy_extract_posterior(variable = adj_dot_unk) %>% #<<
    
    # low confidence interval #<<
    mutate(adj_low_unk=future_pmap(
      .l = select(.,
                  positive_number_test=total_low_round,
                  total_number_test=total_den_low_round),
      .f = possibly(serosvy_unknown_sample_posterior,otherwise = NA_real_), 
      true_positive = true_positive,
      false_negative = false_negative,
      false_positive = false_positive,
      true_negative = true_negative)) %>%
    serosvy_extract_posterior(variable = adj_low_unk) %>%
    
    # upper confidence interval #<<
    mutate(adj_upp_unk=future_pmap(
      .l = select(.,
                  positive_number_test=total_upp_round,
                  total_number_test=total_den_upp_round),
      .f = possibly(serosvy_unknown_sample_posterior,otherwise = NA_real_),
      true_positive = true_positive,
      false_negative = false_negative,
      false_positive = false_positive,
      true_negative = true_negative)) %>%
    serosvy_extract_posterior(variable = adj_upp_unk) %>% 
    
    # union all outputs
    union_all(out)
  
  # print the progress! (recommended)
  out %>% print() #<<
  
}

toc()

outcome_01_adj <- out  %>% #<<
  mutate(rowname=as.numeric(rowname)) %>% 
  arrange(rowname) %>% 
  select(-rowname) 
```


<!-- --- -->

<!-- # Results -->

<!-- -- -->

<!-- - For the __survey__ side, we provide functions to calculate single prevalences and generate __tidy outputs__ for a hierarchical bayesian approach that incorporates the __unknown__ test performance -->

<!-- -- -->

<!-- - For the __serology__ side, we gather functions of methods available in GitHub repos to solve the inherited misclassification in population-based serological surveys. -->

<!-- -- -->

<!-- - We created a __workflow__ that simulates the usage of an __imperfect test__ to measure an outcome within a __multi-stage sampling__ survey design and incorporate the test uncertainty into the sampling design uncertainty. -->

<!-- -- -->

<!-- - The `serosurvey` R package was used in the analysis code of the Lima seroprevalence study.  -->

<!-- -- -->

<!-- [1] Reyes-Vega, Mary F., et al. "SARS-CoV-2 prevalence associated to low socioeconomic status and overcrowding in an LMIC megacity: A population-based seroepidemiological survey in Lima, Peru." _EClinicalMedicine_ 34 (2021): 100801. https://doi.org/10.1016/j.eclinm.2021.100801 -->

---

# Take home messages

--

- The `serosurvey` R package facilitated the estimation of prevalence in the __current epidemic__.

--

- The `serosurvey` R package wanted to reduce the __reproducibility gap__ of serological survey analysis using diagnostic tests with unknown performance within a free software environment.

--

- The usage of an __R package plus a workflow__ may improve our preparedness and response for next epidemics.

--

- Specially in our capacity to __generate comparable results__ from studies with similar epidemiological designs and statistical analysis plans.

--

- The `serosurvey` R package was used in the analysis code of the Lima seroprevalence study. 

--

[1] Reyes-Vega, Mary F., et al. "SARS-CoV-2 prevalence associated to low socioeconomic status and overcrowding in an LMIC megacity: A population-based seroepidemiological survey in Lima, Peru." _EClinicalMedicine_ 34 (2021): 100801. https://doi.org/10.1016/j.eclinm.2021.100801

---

# Next steps

--

- Adjust to the __rOpenSci__ packaging guidelines

--

- Submit package to __CRAN__

--

- Integrate it to the R Epidemic Consortium __RECON__

--

- Welcome new functions and workflows to the package via __GitHub__!

--

.footnote[
[1] https://devguide.ropensci.org/building.html

[2] https://www.repidemicsconsortium.org/resources/guidelines/

[3] https://twitter.com/avallecam/status/1312670190694535169
]



---

class: center, middle

# Thanks for your attention! <br/> Happy to chat with you!

## <br/>Andree Valle Campos<br/><br/> `r icons::fontawesome$brands$twitter` `r icons::fontawesome$brands$github` __`@avallecam`__ <br/> `r icons::fontawesome$solid$inbox` __avallecam@gmail.com__

